\section{Introduction}\label{sec:introduction}

Regular expressions were introduced by Stephen Kleene in the 1950s as a notation for
describing the patterns that a finite-state machine can recognise. The idea was to
capture, with a few simple operators for choice, concatenation and repetition,
exactly the languages accepted by finite automata, and that correspondence between a
compact piece of syntax and a machine has proved to be one of the most durable ideas
in computer science. Regular expressions are interesting precisely because they sit
at this meeting point of theory and practice. On the theoretical side they describe a
well-understood class of languages and come with a rich algebra. On the practical
side they are
everywhere, from the lexical analysers inside every compiler, through search and
text processing, to validation of input and scanning of network traffic. A notation
that a programmer types in seconds is backed by decades of automata theory, and that
is what makes getting its foundations right worthwhile.

The usual way to run a regular expression is to compile it into a finite automaton
and then let the machine consume the input, using one of the classical
constructions of Thompson~\cite{Thompson1968}, Glushkov~\cite{Glushkov1962}, or
McNaughton and Yamada~\cite{McNaughtonYamada1960}. There is also an older and more
direct idea, due to Brzozowski~\cite{B62}, that this thesis takes as its starting
point. Rather than building a separate machine, one can differentiate the expression
itself: the derivative of an expression with respect to a symbol describes what
remains to be matched after that symbol has been consumed, so iterating it across an
input word is already a way of running the expression. When derivatives are kept in
a canonical form there are only finitely many of them, and taking those canonical
derivatives as states yields a deterministic automaton with no auxiliary
construction at all. The approach has aged well and was reexamined from a modern
implementation perspective by Owens, Reppy and Turon~\cite{OwensReppyTuron}.

Exact matching, however, is often stricter than an application really wants. Text is
noisy, symbols are mistyped or substituted, and two inputs that a person would treat
as the same can differ in a handful of places. This is why fuzziness is interesting:
a great deal of real matching is approximate, and a theory that insists on equality
of symbols cannot describe it. A natural way to relax the rigidity is to replace
equality between symbols by a graded notion of similarity, so that two symbols match
not only when they are identical but also when they are close enough, with closeness
measured by a value in $[0,1]$ and combined using triangular norms and the machinery
of fuzzy logic~\cite{met05}. The same instinct drives similarity-based logic
programming~\cite{Sessa2002}, where unification is loosened into matching up to a
similarity relation. The question this thesis asks is what becomes of Brzozowski's
derivative theory under this relaxation: if the consumed symbol need only be similar
to the first symbol of a word in the language, up to a threshold $\mu$, does the
clean derivative-based account of regular languages and automata still hold together?

The answer is that it does, and for a reason that is simple once it is isolated.
When similarity is combined with the minimum triangular norm, cutting the similarity
relation at a threshold $\mu$ produces an equivalence relation, and a fuzzy
derivative over the original alphabet is then nothing more than an ordinary
derivative over the quotient alphabet obtained by identifying symbols that are
similar within $\mu$. Through this quotient-alphabet construction the crisp theory
transfers to the fuzzy setting almost wholesale, so regularity, the correspondence
between the syntactic and the semantic derivative, finiteness of the set of
derivatives, and correctness of the derivative automaton all carry over, and the
resulting fuzzy automaton accepts precisely the words that lie within tolerance
$\mu$ of some word of the original language.

Stating such results is one thing; being sure of them is another, and this is where
formalisation enters and why it is arguably the most important part of the thesis.
Derivative arguments are easy to get subtly wrong, and the finiteness claims that
make the automaton construction terminate are exactly the kind of constructive
statement where informal reasoning tends to gloss over real difficulty. A proof
assistant removes that doubt: it forces every simplification, every side condition
and every induction to be spelled out, and it checks the result mechanically, so a
theorem that has been formalised is one whose proof has been audited down to the last
step by a program that cannot be talked into accepting a gap. Formalisation is
interesting in its own right because it turns mathematical confidence from a matter
of careful reading into a matter of machine verification, and it is especially
valuable for a theory like this one that is meant to be both correct and executable.
For that reason the crisp theory is not only proved on paper but also mechanised in
the Rocq proof assistant~\cite{Rocq}, building on the Coquand--Siles formalisation of
regular-expression equivalence via Brzozowski
derivatives~\cite{CoquandRegexEquality,RocqRegexpBrzozowski}. The mechanisation
covers the syntax and semantics of regular expressions, the nullability test,
canonicalisation and its correctness, the derivative correspondence and the
correctness of the derivative automaton, and it supplies the similarity
infrastructure on which the fuzzy layer rests. The full development is available as
an accompanying repository~\cite{OurRocqRepo}, and rather than confine it to a
section of its own the code is presented next to the mathematics it certifies.

The rest of the thesis is organised as follows. Section~\ref{sec:literature} places
the work in the context of the existing literature on regular expressions,
derivatives, their formalisation, and fuzzy automata.
Section~\ref{sec:methodology} contains the technical development in two layers: the
crisp theory of derivatives and the derivative automaton, presented together with
its Rocq formalisation, followed by the fuzzy layer with its definitions, the
quotient construction, and the fuzzy automaton. Section~\ref{sec:conclusion}
summarises what has been achieved and points to the directions that follow from it,
chief among them completing the mechanisation of the fuzzy layer.
